--- 
title: "PL 9239 Introduction to Data Science for Politics and IR"
author: "Christian Arnold, Cardiff University"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [IntroDS.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This is the website for the Intro to Data Science Class."
---




# Welcome! {-}
Welcome to this module. This module teaches you how to make the most of data. You will learn to handle and analyse data efficiently and how to communicate results effectively, e.g. by visualising them. All we do in this class will revolve around a key theme: How can we use the information in data to study politics and international relations?


## Data to Answer Your Questions {-}
<!-- Why are we doing this? -->

Today, data are everywhere. Data are not only collected for scientific studies. Governments, companies and non-profit organisations are amassing and increasingly sharing data at an unprecedented level. 

Data on their own can actually be a bit boring. Few people might enjoy just looking at a string of numbers. The whole thing changes, however, if we can use this data to answer a question we are interested in. 


Imagine you are working for an NGO that helps implement voting observation missions. In the run-up for an election in a fragile democracy, your NGO is invited to help deploy voting observation missions. Even though you are partnering up with local civil society organisations, you cannot send election observers too all polling stations. Luckily, this is not strictly necessary: As long as you properly randomise the polling stations you want to observe, you can infer the real, countrywide numbers on the basis of just a few hundred few hundred observations. In this class, we will learn why this is working and how you can do inference yourself. 


Or take another example: A new law introduced mandatory environmental protection for farm land. A year later, biologists measure that harmful substances in ground water are down by 20%. Was this the effect of the policy? Or was it maybe just a particularly rainy year that washed all pollutants away? Together we will learn how we can find answers by asking our data the right kind of questions.


Together we will build the knowledge and skills necessary for working with data yourself. The module is particularly relevant if you are planning to undertake a dissertation, since it will help you find the right data and how to work with it. Of course, data literacy is also a highly relevant skills that will contribute to your employability.





## Your Weekly Data Workout{-}
<!-- How to use this material? -->
In this module, we will expose you to core concepts _prior_ to the actual class. This frees up large chunks of time during the class that we can spend on activities where you typically need the most help, such as application of basic material and engaging in deeper discussions and creative work with it. A year ago, all this would have been quite revolutionary. But now, 12 months into this pandemic, this approach will already feel much more intuitive. 

Let me introduce all of our activities one by one.


### 1 Work Through This Homepage {-}

Your first port of call in every week will be the respective chapter in this module homepage. This site is a mixture between a website and a teaching book. It will walk you through the content of each week and you can find the most important content explained here. The website allows you to take your own pace and revisit the parts that you find most challenging as many times as you want and at the pace that you feel comfortable with.

The respective content of the week will be uploaded on Monday morning, latest. I expect that you work through that home page on your own over the course of the week. In case you have questions, please do reach out to me via [email](mailto:arnoldc6@cardiff.ac.uk) or book time in the [office hour](https://tinyurl.com/yayvd5no). We will also use the module cafe every second week for answering any of your questions.



### 2 Apply Your Knowledge in Our Workshops {-}

While the homepage is focusing on knowledge acquisition, our workshops on Tuesdays between 12:00 and 2:00 o'clock will focus on skills development. Learning how to do data analysis requires practice. And practice means failing a lot and learning from the own mistakes. The workshops on Tuesdays are meant to serve as a place where we can jointly explore how to properly manage, analyse and visualise data.

The workshops will always cover the content of the past week, i.e. in the workshop in week 2 we will be covering the content of the homepage from week 1. Please make sure that you come prepared. Workshops will run from week 1 until week 11. 

Over the course of these two hours, you will be working online in small groups. For each week, I will prepare a lab sheet that you will jointly solve. Do not worry in case you struggle: I will be always around to help.

Without participating in these workshops, the assignments will be really challenging. I strongly recommend you do not miss a single session.



### 3 Bring Your Questions to the Module Cafés {-}

We will have bi-weekly module cafes where you can discuss any additional questions that you might have. In addition, these module cafes will also serve to answer any question regarding the assignments.

### 4 Read Guideance on Planning and Structuring {-}
To guide you through all the hoops, I will also stay in touch with you with am email every Friday to make sure that you are on board. It is vital that you check your email regularly, since we will only communicate with you through your university email. 


## The Module Webpage {-}
<!-- What is this page for? -->
This website you are visiting will be the main learning tool for this term's _Introduction to Data Science for Politics and IR_. Let us take a look how this works. 

Obviously, I will guide you through the material on this web-page via text. Think of it as a replacement for the lecture slides that you might find in other contexts: Here, you can find the core concepts and central ideas of this class. Since analysing data requires a solid understanding, you can find a fair amount of maths here, like so $$ \theta \sim N(0,1)$$


But not all is text. Sometimes, nothing beats someone actually explaining you core ideas. For that purpose, I will record small videos and embedd them here. The best thing about videos: You can watch them as many times as you want...


```{r, eval=knitr::is_html_output(excludes = "epub"), results = 'asis', echo = F}
cat(
'<iframe width="560" height="315" src="https://www.youtube.com/embed/1QJ6P8Yxil4" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>'
)
```

Finally, this format is also really handy, because we will be able to take a look at R code. R is a computer language invented for data analysis, and you will learn how to use it to do data analysis. This is how it looks like when we will be covering R code.
```{r, echo = TRUE}
# All text in a line after the hashtag is a comment.
# Here we assign some variables
a <- 1 
b <- 2
# This is how to add them and assign the result to another variable
c <- a+b
# And finally we print the result
cat(c)
```





Last but not least, each week will have some readings that will reflect the core content of the respective week from yet another perspective. I will add pointers to the readings here on the homepage whenever appropriate.


## Getting in Touch {-}
If you have questions or need any kind of help, feel free to get in touch. 

* Our weekly workshops and our bi-weekly module café are explicitly made so that you can ask questions.
* You can always drop me an [email](mailto:arnoldc6@cardiff.ac.uk). I will respond at the end of the working day. 
* You can also [book time](https://tinyurl.com/yayvd5no) in my office hours 
  * Tuesday 3-4pm;
  * Wednesday 3.15-4.15pm;
  * Friday 5.15-6.15pm.



<!-- To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.org/tinytex/>. -->

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


```{r echo=FALSE}
# -- housekeeping for the rest of the class ------------------------------------ 

# Libraries
library(foreign)
library(xtable)


# data: Class surveys
dat18 <- read.csv("preparing_data/data/class_survey/lecture_survey_18.csv")
dat19 <- read.csv("preparing_data/data/class_survey/lecture_survey_19.csv")
dat <- read.csv("preparing_data/data/class_survey/lecture_survey_21.csv")

# names(dat)
attach(dat)

# Predefined Cardiff colours
# primary
cardiffred <- rgb(211,55,74, maxColorValue = 255)
cardiffblack <- rgb(35,31,32, maxColorValue = 255)
cardiffgrey <- rgb(47,68,78, maxColorValue = 255)
cardiffgold <- rgb(189,158,94, maxColorValue = 255)
# secondary 
cardiffblue <- rgb(21,44,81, maxColorValue = 255)
cardiffpurple1 <- rgb(29,15,51, maxColorValue = 255)
cardiffpurple2 <- rgb(60,44,89, maxColorValue = 255)

```





